Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2020-05-19 Manuscript/manuscriptPGPE.tex
Add first POPG-ECC results and update figure and text

\section{Methods}
\subsection{Experimental C--H bond order parameters}
The headgroup and glycerol backbone C--H bond order parameter magnitudes and signs of POPE and POPG
were determined by measuring the chemical-shift resolved dipolar splittings
with a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04} and
%The corresponding order parameter signs were measured with a
S-DROSS experiments~\cite{gross97} using natural abundance $^{13}$C solid state NMR spectroscopy
as described previously \cite{ferreira13,ferreira16}.
%The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
%Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
%The following experimental setups were used.
POPE and POPG powder were purchased from Avanti polar lipids.
The NMR experiments were identical to our previous work~\cite{antila19}.
\todo{Is this enough and correct, or should we repeat some methods from the NMRlipidsIVps paper?}

@@

\subsection{Headgroup and glycerol backbone order parameters of POPE and POPG from $^{13}$C NMR}
Absolute values of the headgroup and glycerol backbone order parameters from PE and PG lipids are measured
previously using $^2$H NMR~\cite{seelig76,gally81,wohlgemuth80,borle85}. Because also the order parameter
signs bear essential information about the lipid structures \cite{botan15,ollila16}, we measured the
magnitudes and signs of POPE and POPG C--H bond headgroup and glycerol backbone order parameter in liquid phase
using the 
%order parameters The order parameters for the glycerol backbone and headgroup s were determined
2D-RPDLF and S-DROSS experiments, as described previously \cite{ferreira13,ferreira16,antila19}.
For POPE, the glycerol backbone and $\alpha$-carbon peaks in INEPT spectra were assigned based on
previously measured POPC spectra~\cite{ferreira13} and
the $\beta$-carbon peak was assigned based on $^{13}$C chemical shift table for amines available
at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm} (Fig. \ref{POPEspectra}).
For POPG, the glycerol backbone peaks in INEPT spectra were assigned based on
previously measured POPC spectra~\cite{ferreira13}, while $\alpha$ and  $\gamma$-carbon peaks
\todo{How were these assigned?} (Fig. \ref{POPGspectra}). The numerical value of the $\beta$-carbon
order parameter could not be determined, because its peak overlapped with the g$_2$ peak from glycerol backbone in POPG.
However, the order parameter of $\beta$-carbon is expected to be clearly smaller than for g$_2$
based on previous $^2$H NMR measurements \cite{wohlgemuth80,gally81,borle85}.
Therefore, the beginning of the S-DROSS curve gives the sign for g$_2$ order parameter and end for $\beta$ (Fig. \ref{POPGspectra} (E)).
This is confirmed with SIMPSON calculations using negative value for g$_2$ and positive value for $\beta$ order parameter (Fig. \ref{POPGsimpson}).
The POPE experiments were recorded at 310~K and POPG experiments at 298~K, where the bilayers are in the liquid disordered phase \cite{marsh13}.
\todo{Details to be checked by Tiago}.

@@

\clearpage
\subsection{Calcium binding to PE and PG lipid bilayers}
The headgroup order parameters of PC lipids can used to measure ion binding
affinity to lipid bilayers, because their magnitude is linearly proportional
to the amount of bound charge in bilayer according to the molecular
electrometer concept \cite{seelig87,catte16}. The molecular electrometer concept
can be used also for bilayers containing PC lipids mixed with charged
lipids \cite{borle85,macdonald87,roux90,antila19}.
The electrometer concept has been very useful in evaluating ion binding affinity in simulations
against experiments, because the headgroup order parameter changes as a function of ion concentration
can be directly compared with experiments \cite{catte16,melcr18,antila19}.

Calcium binding affinity to PC and PS lipid bilayers was not correctly described by
any of the standard MD simulation forced fields \cite{catte16,antila19}, while
recently introduced force field with electronic continuum correction (ECC) performed better \cite{melcr18}.
The decrease of $\alpha$-carbon order parameter of PC lipids in PC:PG mixtures as a function
of calcium concentration is close to experiments CHARMM36 simulations (Fig. \ref{changesWITHCaClPG}),
but the decrease of $\beta$-carbon order parameter seems to be overestimated.
However, the $\beta$-carbon order parameter was not actually measured from these samples,
but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
On the other hand, the simulations are not converged.
\todo{We still need more data to finish the discussion. More detailed discussion is in https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/12}
% The result is similar to the $\sim$200~ns simulations with PC lipids in previous work \cite{catte16}.
%However, when simulation was continued for $\mu$s, the binding affinity substantially increased
%and interpretation was that calcium overbinds to PC lipid in CHARMM36. Therefore, the
%conclusion seems to be similar here, although the new NBfix parameters may complicate
%the situation \todo{The status of NBfix parameters in these simulations should be checked.}.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithCaClPG.eps}
  \caption{\label{changesWITHCaClPG}
    (left) The headgroup order parameters of PC from PC:PG mixtures as a function CaCl$_2$
    concentration from experiments \cite{borle85,macdonald87} and CHARMM36 simulations.
    Note that beta order parameter is calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}, not actually measured.
    (right) The headgroup order parameters of PG from PC:PG mixtures as a function CaCl$_2$
    concentration from experiments \cite{borle85} and CHARMM36 simulations.
  }
\end{figure*}

@@

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CAdensPG.eps}
  \caption{\label{CAdensPG}
    Calcium ion density profiles along membrane normal
    from different simulations with PG lipids.
  }
\end{figure}

>>>>

\section{Methods}
\subsection{Experimental C--H bond order parameters}
The headgroup and glycerol backbone C--H bond order parameter magnitudes and signs of POPE and POPG
were determined by measuring the chemical-shift resolved dipolar splittings
with a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04} and
%The corresponding order parameter signs were measured with a
S-DROSS experiments~\cite{gross97} using natural abundance $^{13}$C solid state NMR spectroscopy
as described previously \cite{ferreira13,ferreira16}.
%The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
%Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
%The following experimental setups were used.
POPE and POPG powder were purchased from Avanti polar lipids.
The NMR experiments were identical to our previous work~\cite{antila19}.
\todo{Is this enough and correct, or should we repeat some methods from the NMRlipidsIVps paper?}
The POPE experiments were recorded at 310~K and POPG experiments at 298~K, where the bilayers are in the liquid disordered phase \cite{marsh13}.

@@

\subsection{Headgroup and glycerol backbone order parameters of POPE and POPG from $^{13}$C NMR}
Absolute values of the headgroup and glycerol backbone order parameters from PE and PG lipids are measured
previously using $^2$H NMR~\cite{seelig76,gally81,wohlgemuth80,borle85}. Because also the order parameter
signs bear essential information about the lipid structures \cite{botan15,ollila16}, we measured the
magnitudes and signs of POPE and POPG C--H bond headgroup and glycerol backbone order parameter in liquid phase
using the 
%order parameters The order parameters for the glycerol backbone and headgroup s were determined
2D-RPDLF and S-DROSS experiments, as described previously \cite{ferreira13,ferreira16,antila19}.
For POPE, the glycerol backbone and $\alpha$-carbon peaks in INEPT spectra were assigned based on
previously measured POPC spectra~\cite{ferreira13} and
the $\beta$-carbon peak was assigned based on $^{13}$C chemical shift table for amines available
at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm} (Fig. \ref{POPEspectra}).
For POPG, the glycerol backbone peaks in INEPT spectra were assigned based on
previously measured POPC spectra~\cite{ferreira13}, while $\alpha$ and  $\gamma$-carbon peaks
\todo{How were these assigned?} (Fig. \ref{POPGspectra}). The numerical value of the $\beta$-carbon
order parameter could not be determined, because its peak overlapped with the g$_2$ peak from glycerol backbone in POPG.
However, the order parameter of $\beta$-carbon is expected to be clearly smaller than for g$_2$
based on previous $^2$H NMR measurements \cite{wohlgemuth80,gally81,borle85}.
Therefore, the beginning of the S-DROSS curve gives the sign for g$_2$ order parameter and end for $\beta$ (Fig. \ref{POPGspectra} (E)).
This is confirmed with SIMPSON calculations using negative value for g$_2$ and positive value for $\beta$ order parameter (Fig. \ref{POPGsimpson}).
\todo{Details to be checked by Tiago}.

@@

\clearpage
\subsection{Calcium binding to PE and PG lipid bilayers}
\begin{figure*}[b]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithCaClPG1PC1.eps}
  \caption{\label{changesWITHCaClPG}
    Modulation of headgroup order parameters of POPC ({\it left}) and POPG ({\it right}) in POPC:POPG (1:1)
    mixture upon addition of CaCl$_2$ in 298 K temperature from experiments \cite{borle85,macdonald87} and simulations.
    The $\beta$-carbon order parameter of POPC (dashed line on top left) is not directly measured but
    calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
    The changes with respect to the systems without CaCl$_2$ are shown for other data than
    for the $\alpha$-carbon of POPG for which experimental order parameter is not available.
    %(right) The headgroup order parameters of PG from PC:PG mixtures as a function CaCl$_2$
    %concentration from experiments \cite{borle85} and CHARMM36 simulations.
  }
    \todo{CHARMM36 simulations should be longer and with Na couterions.}
\end{figure*}

To evaluate calcium binding to PG containing lipid bilayers in simulations,
we calculated the changes in headgroup order parameters upon addition of CaCl$_2$
to POPC:POPG (1:1) and (4:1) mixtures for which experimental data is available \cite{borle85,macdonald87}.
The headgroup order parameters of PC lipids can be used to measure the ion binding
affinity to lipid bilayers and evaluate MD simulations against experimental data,
because their magnitude is linearly proportional to the amount of bound charge in bilayer
and order parameters can be directly calculated from MD simulations \cite{seelig87,catte16}.
This molecular electrometer concept can be used also for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90,antila19}.
%The electrometer concept has been very useful in evaluating ion binding affinity in simulations
%against experiments, because the headgroup order parameter changes as a function of ion concentration
%can be directly compared with experiments \cite{catte16,melcr18,antila19}.
%

%Calcium binding affinity to PC and PS lipid bilayers was not correctly described by
%any of the standard MD simulation forced fields \cite{catte16,antila19}, while
%recently introduced force field with electronic continuum correction (ECC) performed better \cite{melcr18}.
As previously observed for PC and PS lipids, the decrease in PC headgroup order parameters with
increasing CaCl$_2$ concentration is overestimated in Slipids and Lipid17 simulations (Figs. \ref{changesWITHCaClPG} and \ref{changesWITHCaClPG1PC4})
indicating too strong binding affinity of calcium into the bilayers.
\todo{CHARMM results to be mentioned once we have the new simulations.}
%The decrease of $\alpha$-carbon order parameter of PC lipids in PC:PG mixtures as a function
%of calcium concentration is close to experiments CHARMM36 simulations (Fig. \ref{changesWITHCaClPG}),
%but the decrease of $\beta$-carbon order parameter seems to be overestimated.
%However, the $\beta$-carbon order parameter was not actually measured from these samples,
%but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
%On the other hand, the simulations are not converged.
The behaviour of $\beta$-carbon order parameter of POPG is similar to POPC upon addition of calcium in experiments,
exhibiting a rapid decrease with CaCl$_2$ concentrations below 10 mM,
and then a more modest decrease with larger concentrations (Fig. \ref{changesWITHCaClPG}) \cite{borle85}.
The behaviour is qualitatively different than in the $\beta$-carbon of PS lipids which increase upon addition of CaCl$_2$.
Lipid17 and Slipids force fields correctly capture the PG $\beta$-carbon order parameter response to CaCl$_2$
even thought the binding affinity was too large based on the comparison of PC headgroup order parameter changes with experiments.
\todo{To be finished once we have the new CHARMM simulations.} For the response of PG $\alpha$-carbon order parameters to CaCl$_2$,
different force fields give different predictions, but experimental data to evaluate these predictions is not available. 
% The result is similar to the $\sim$200~ns simulations with PC lipids in previous work \cite{catte16}.
%However, when simulation was continued for $\mu$s, the binding affinity substantially increased
%and interpretation was that calcium overbinds to PC lipid in CHARMM36. Therefore, the
%conclusion seems to be similar here, although the new NBfix parameters may complicate
%the situation \todo{The status of NBfix parameters in these simulations should be checked.}.

@@

\todo{We still need more data to finish the discussion. More detailed discussion is in https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/12}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CAdensPG1PC1.eps}
  \caption{\label{CAdensPG}
    Calcium ion density profiles along membrane normal from simulations of POPC:POPG (1:1) mixtures with different force fields.
  }
\end{figure}

### 2020-05-19 Manuscript/manuscriptPGPEsuppl.tex
Add first POPG-ECC results and update figure and text

\bibliography{refs.bib}

>>>>

\section{Calcium binding to POPC:POPG (4:1) mixtures}
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithCaClPG1PC4.eps}
  \caption{\label{changesWITHCaClPG1PC4}
    Modulation of headgroup order parameters of POPC ({\it left}) and POPG ({\it right}) in POPC:POPG (4:1)
    mixture upon addition of CaCl$_2$ in 298 K temperature from experiments \cite{macdonald87} and simulations.
    The changes with respect to the systems without CaCl$_2$ are shown for other data than
    for the $\alpha$-carbon of POPG for which experimental order parameter is not available.
  }
  \todo{CHARMM36 simulations should be longer and with Na couterions.}
\end{figure*}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CAdensPG1PC4.eps}
  \caption{\label{CAdensPG1PC4}
    Calcium ion density profiles along membrane normal from simulations of POPC:POPG (4:1) mixtures with different force fields.
  }
\end{figure}

\clearpage
\bibliography{refs.bib}
